Next Generation Planning -- Characterizing Exposure to and Shared Knowledge on Drivers of Ecological Change in the St. Lawrence System

### 2020-04-18 Revisions2/Beauchesne-eDriversMS.Rmd
prepare 2nd round of review and modify figure 1

---
title: Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence System
author: David Beauchesne^1,2,\*^, Rémi M. Daigle^2^, Steve Vissault^3^, Dominique Gravel^3^, Andréane Bastien^4^, Simon Bélanger^5^, Pascal Bernatchez^5^, Marjolaine Blais ^6^, Hugo Bourdages^6^, Clément Chion^7^, Peter S. Galbraith^6^, Benjamin Halpern^8,9^, Camille Lavoie^2^, Christopher W. McKindsey^6^, Alfonso Mucci^10^, Simon Pineault^11^, Michel Starr^6^, Anne-Sophie Ste-Marie^4^, Philippe Archambault^2^
fontsize: 12pt
font-family: Times New Roman
output:
  pdf_document:
    toc: false
    number_sections: true
    keep_tex: true
  word_document:
    toc: false
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{listings}
   - \usepackage[svgnames]{xcolor}
   - \usepackage{lscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{bbm}
   - \usepackage{float}
   - \usepackage[normalem]{ulem}
   - \definecolor{blue}{HTML}{20cad5}
relativeurls: true
bibliography: Beauchesne-eDriversMS.bib
csl: frontiers.csl
link-citations: yes
---

<!--
rmarkdown::render('./Revisions/Beauchesne-eDriversMS.Rmd', 'pdf_document')
rmarkdown::render('./eDriversMS-v2.rmd', 'word_document')
-->

@@

The St. Lawrence is a vast and complex socio-ecological system providing a
wealth of services sustaining numerous economic sectors. These ecosystems are
subject to significant human pressures that overlap and potentially interact
with climate driven environmental changes. Our objective in this paper is to
systematically characterize the distribution and intensity of drivers \textcolor{blue}{of change} in the
St. Lawrence System. To do so, we launch *eDrivers*, an open knowledge
platform gathering experts committed to structuring, standardizing and sharing
knowledge on drivers \textcolor{blue}{of change} in support of science and management. We gathered data on
22 coastal, climate, fisheries and marine traffic drivers through
collaborations, existing environmental initiatives and open data portals. We
show that few areas of the St. Lawrence are free of cumulative exposure. The
Estuary, the Anticosti Gyre and coastal areas are particularly exposed,
especially in the vicinity of urban centers. We identified 6 areas of distinct
cumulative exposure regime that show that certain drivers \textcolor{blue}{of change} typically co-occur
in different regions of the St. Lawrence and that coastal areas are exposed to
all driver types. Of particular concern are two threat complexes capturing
most exposure hotspots that show the convergence of contrasting exposure
regimes at the head of the Laurentian Channel. *eDrivers* was built on a
series of guiding principles upholding existing data management and open
science standards. We therefore expect it to evolve through time to address
knowledge gaps and refine current driver layers. Ultimately, we believe that
*eDrivers* represents a much needed solution that could radically influence
broad scale research and management practices by increasing knowledge accessibility
and interoperability.

@@

As elsewhere in the world [see @halpern2015a], this intensifying human footprint
will likely result in increasingly intricate environmental exposure regimes,
*i.e.* suites of overlapping \textcolor{red}{\sout{and potentially interacting}} environmental drivers \textcolor{blue}{of change}
threatening ecosystems, habitats or ecological communities. Drivers \textcolor{blue}{of change}, often referred to as stressors or pressures \textcolor{blue}{and hereafter called drivers for simplicity}, are any externalities that affect
environmental processes and disturb natural systems. Drivers may originate from
natural or human-induced biophysical processes (*e.g.* sea surface\textcolor{blue}{-water} temperature
anomalies and hypoxia) or directly from anthropogenic activities (*e.g.*
fisheries and marine pollution). The potential for complex interactions between
\textcolor{blue}{co-occurring} drivers is the largest uncertainty when studying or predicting environmental impacts [@darling2008; @cote2016]. Multiple drivers can combine non-linearly and result in effects that are greater (synergistic effect) or lower (antagonistic effect) than the sum of individual effects
[@crain2008; @darling2008; @cote2016].

@@

There are now initiatives that \textcolor{blue}{facilitate the data gathering process} \textcolor{red}{\sout{address this issue}} by assembling, organizing and sharing environmental knowledge, such as the Ocean Biogeographic Information
System [OBIS; @obis2019] for biotic data and Bio-ORACLE [@tyberghein2012] for
abiotic data.
However, equivalent platforms for environmental drivers have largely focused on
single drivers (e.g. Global Fishing Watch) and platforms collating
data and knowledge on multiple drivers in a comparable and interoperable way
remain conspicuously missing [but see @halpern2015]. This is in spite of
integrated management and assessment approaches needing efficient data reporting,
standardized data management practices and tools tailored to the study of the
effects of multiple drivers [@dafforn2016; @stock2018]. An additional objective
thus emerged in the process of addressing our initial goals: \textcolor{red}{\sout{sharing the knowledge gathered through the description of drivers in the St. Lawrence.}} \textcolor{blue}{sharing information about the distribution and intensity of drivers of change in the St.Lawrence.}

Our main objective in this study is to characterize the distribution and
intensity of drivers in the St. Lawrence. More specifically, we aim to: 1)
identify areas of high cumulative exposure\textcolor{red}{\sout{, 2) identify drivers that are likely to interact in the St. Lawrence}} and \textcolor{red}{\sout{3)}} \textcolor{blue}{2)} characterize areas with similar cumulative exposure regimes.
\textcolor{red}{\sout{Here we focus solely on drivers, one of four elements required to evaluate environmental impacts, *i.e.* drivers, ecosystem components of interest (*e.g.* habitats or species) and the exposure and the vulnerability of the ecosystem components to drivers.}}
\textcolor{blue}{Here, we focus solely on drivers, aiming at better understanding their distribution in the St. Lawrence and at facilitating the widespread availability of their characterization for holistic assessments and management approaches.}
We achieve these objectives with the development of an
open knowledge platform named *eDrivers*. The platform was designed to
facilitate collaboration, real-time assessments of cumulative exposure and to
evolve with the addition of information and threats to the St-Lawrence ecosystems.

@@

## Estuary and Gulf of St. Lawrence

The St. Lawrence System is formed by the Estuary and the Gulf of St. Lawrence
(Figure \ref{egsl}). The Estuary is defined by the seawater limit
close to Îles d'Orléans to the west and by its connexion to the Gulf
near Pointe-des-Monts.
Estuarine circulation characterizes the Estuary. The surface layer is
composed of a freshwater seaward outflow primarily from the Great Lakes Basin
and through the St. Lawrence River, upstream of the Estuary. Atlantic waters are
flowing mainly from the Cabot Strait landward through the Laurentian Channel
(see below).

@@

\textcolor{red}{\sout{The list of drivers for which we sought data was informed by}} \textcolor{blue}{Drivers selection was informed by} a global
cumulative impact assessment initiative [@halpern2008a; @halpern2015a] \textcolor{blue}{and available from the National Center for Ecological Analysis and Synthesis (NCEAS) online data repository (Halpern et al. 2015)},
regional holistic evaluations of the state of the St. Lawrence [@dufour2007;
@benoit2012], and communications with regional experts (Table 1).
\textcolor{red}{\sout{We integrated datasets from regional experts and also use global data from the global cumulative impact assessment initiative [@halpern2008a; @halpern2015a]  available from the National Center for Ecological Analysis and Synthesis (NCEAS) online data repository [Table 1; @halpern2015].}} We selected global data that were
unavailable at the regional scale and that were available at a resolution
adequate for use at the scale of the St. Lawrence (*e.g.* marine pollution).

We characterized the intensity and distribution of 22 drivers
(Table 1;). Drivers incorporated in the analyses are varied in origin,
*i.e.* from terrestrial (*e.g.* nutrient input) to marine (*e.g.* shipping),
and from large scale biophysical processes (*e.g.* temperature anomalies) to
localized anthropogenic activities (*e.g.* fisheries). Drivers were divided
into 4 groups: coastal, climate, fisheries and marine traffic (Table 1). All
data layers and methodologies are described in the supplementary materials.

Drivers with non-normal frequency distributions were log-transformed,
\textcolor{blue}{as in Halpern et al. 2008, to minimize the impact of extreme outliers that may or may not be real observations} (Figure S1)\textcolor{blue}{,} and all drivers were scaled between 0 and 1 to allow driver comparisons. The 99th quantile of individual driver distribution was used as the upper bound for scaling \textcolor{red}{\sout{to control for extreme values and produce maps of individual drivers (Figure S2)}}. All drivers were embedded in a regular grid composed of 245604 $1 km^2$ hexagonal cells to construct the integrated dataset used for the
analyses \textcolor{blue}{(Figure S2)}.

@@

Hypoxia is mainly distributed in the Laurentian, Anticosti and Esquiman Channels, with the head of the Channels most exposed (Figure \ref{kernel}A). Demersal destructive fisheries are located along the Laurentian Channel, the heads of the Anticosti and Esquiman Channels and around the Magdalen Islands (Figure \ref{kernel}B). By combining both drivers, we can observe that hypoxia and demersal destructive fisheries co-occur mostly at high relative intensity (Figure \ref{kernel}D) in the vicinity of the Anticosti Gyre and the heads of the Esquiman and Anticosti Channels (Figure \ref{kernel}C, Box 1).

@@

As such, we are launching *eDrivers*, an open knowledge platform focused on
sharing knowledge on the distribution and intensity of drivers and on gathering
a community of experts committed to structuring, standardizing and sharing
knowledge on drivers in support of science and management. In launching this
initiative, our objective is to uphold the highest existing standards of data
management and open science. We identified four guiding principles to meet this
objective and that guide the structure of the platform (Figure \ref{diag}).

@@

Through *eDrivers*, these observations will be iteratively improved towards an increasingly robust assessment of cumulative exposure and threat complexes as gaps in knowledge are addressed or approaches to describe drivers are refined. Arguably, the most meaningful benefit anticipated from *eDrivers* will be the gain in efficient access to comparable data-based knowledge on the exposure of ecosystems to multiple threats. This could pay quick scientific and management dividends by drawing on the knowledge and efforts of a wide range of contributors, by expanding avenues of scientific inquiry, by decreasing overall effort duplication and research costs, and by increasing research efficiency [@franzoni2014].

>>>>

---
title: Next Generation Planning -- Characterizing Exposure to and Shared Knowledge on Drivers of Ecological Change in the St. Lawrence System
author: David Beauchesne^1,2,\*^, Rémi M. Daigle^2^, Steve Vissault^3^, Dominique Gravel^3^, Andréane Bastien^4^, Simon Bélanger^5^, Pascal Bernatchez^5^, Marjolaine Blais ^6^, Hugo Bourdages^6^, Clément Chion^7^, Peter S. Galbraith^6^, Benjamin Halpern^8,9^, Camille Lavoie^2^, Christopher W. McKindsey^6^, Alfonso Mucci^10^, Simon Pineault^11^, Michel Starr^6^, Anne-Sophie Ste-Marie^4^, Philippe Archambault^2^
fontsize: 12pt
font-family: Times New Roman
output:
  pdf_document:
    toc: false
    number_sections: true
    keep_tex: true
  word_document:
    toc: false
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{listings}
   - \usepackage[svgnames]{xcolor}
   - \usepackage{lscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{bbm}
   - \usepackage{float}
   - \usepackage[normalem]{ulem}
   - \definecolor{blue}{HTML}{20cad5}
relativeurls: true
bibliography: Beauchesne-eDriversMS.bib
csl: frontiers.csl
link-citations: yes
---

<!--
rmarkdown::render('./Revisions2/Beauchesne-eDriversMS.Rmd', 'pdf_document')
-->

<!--
To discuss with co-authors

- New article title
- Should we present how the platform is used in the analyses or really keep
  that portion fully separate? If we include it, modify answer to question 2
  comment 1

-->

@@

The St. Lawrence is a vast and complex socio-ecological system providing a
wealth of services sustaining numerous economic sectors. ~~\textcolor{red}{These ecosystems are}~~
\textcolor{blue}{This ecosystem is}
subject to significant human pressures that overlap and potentially interact
with climate driven environmental changes. Our objective in this paper is to
systematically characterize the distribution and intensity of drivers \textcolor{blue}{of change} in the
St. Lawrence System. ~~To do so, we launch *eDrivers*, an open knowledge
platform gathering experts committed to structuring, standardizing and sharing
knowledge on drivers \textcolor{blue}{of change} in support of science and management.~~ We gathered data on
22 coastal, climate, fisheries and marine traffic drivers through
collaborations, existing environmental initiatives and open data portals. We
show that few areas of the St. Lawrence are free of cumulative exposure. The
Estuary, ~~the~~ Anticosti Gyre and coastal areas are particularly exposed,
especially in the vicinity of urban centers. We identified 6 areas of distinct
cumulative exposure regime that show that certain drivers \textcolor{blue}{of change} typically co-occur
in different regions of the St. Lawrence and that coastal areas are exposed to
all driver types. Of particular concern are two threat complexes capturing
most exposure hotspots that show the convergence of contrasting exposure
regimes at the head of the Laurentian Channel.
\textcolor{blue}{Sharing the knowledge acquired emerged as a priority to minimize future
environmental assessment efforts. We thus launch *eDrivers*, an open knowledge
platform gathering experts committed to structuring, standardizing and sharing
knowledge on drivers of change in support of science and management.}
*eDrivers* was built on a
series of guiding principles upholding existing data management and open
science standards. We therefore expect it to evolve through time to address
knowledge gaps and refine current driver layers. Ultimately, we believe that
*eDrivers* represents a much needed solution that could radically influence
broad scale research and management practices by increasing knowledge accessibility
and interoperability.

@@

As elsewhere in the world [see @halpern2015a; @halpern2019], this intensifying human footprint
will likely result in increasingly intricate environmental exposure regimes,
*i.e.* suites of overlapping \textcolor{red}{\sout{and potentially interacting}} environmental drivers \textcolor{blue}{of change}
threatening ecosystems, habitats or ecological communities. Drivers \textcolor{blue}{of change}, often referred to as stressors or pressures \textcolor{blue}{and hereafter called drivers for simplicity}, are any externalities that affect
environmental processes and disturb natural systems. Drivers may originate from
natural or human-induced biophysical processes (*e.g.* sea surface\textcolor{blue}{-water} temperature
anomalies and hypoxia) or directly from anthropogenic activities (*e.g.*
fisheries and marine pollution). The potential for complex interactions between
\textcolor{blue}{co-occurring} drivers is the largest uncertainty when studying or predicting environmental impacts [@darling2008; @cote2016]. Multiple drivers can combine non-linearly and result in effects that are greater (synergistic effect) or lower (antagonistic effect) than the sum of individual effects
[@crain2008; @darling2008; @cote2016].

@@

There are now initiatives that \textcolor{blue}{facilitate the data gathering process} \textcolor{red}{\sout{address this issue}} by assembling, organizing and sharing environmental knowledge, such as the Ocean Biogeographic Information
System [OBIS; @obis2019] for biotic data and Bio-ORACLE [@tyberghein2012] for
abiotic data.
However, equivalent platforms for environmental drivers have largely focused on
single drivers (e.g. Global Fishing Watch) and platforms collating
data and knowledge on multiple drivers in a comparable and interoperable way
remain conspicuously missing [but see @halpern2015]. This is in spite of
integrated management and assessment approaches needing efficient data reporting,
standardized data management practices and tools tailored to the study of the
effects of multiple drivers [@dafforn2016; @stock2018].
~~An additional objective
thus emerged in the process of addressing our initial goals:
sharing information about the
distribution and intensity of drivers of change in the St.Lawrence.~~

Our main ~~objective~~ \textcolor{blue}{goal} in this study is to characterize the distribution and
intensity of drivers in the St. Lawrence. More specifically, ~~we aim~~ \textcolor{blue}{our objectives are} to: 1)
identify areas of high cumulative exposure\textcolor{red}{\sout{, 2) identify drivers that are likely to interact in the St. Lawrence}} and \textcolor{red}{\sout{3)}} \textcolor{blue}{2)} characterize areas with similar cumulative exposure regimes.
\textcolor{red}{\sout{Here we focus solely on drivers, one of four elements required to evaluate environmental impacts, *i.e.* drivers, ecosystem components of interest (*e.g.* habitats or species) and the exposure and the vulnerability of the ecosystem components to drivers.}}
\textcolor{blue}{Here, we focus solely on drivers, aiming at better understanding their distribution in the St. Lawrence and at facilitating the widespread availability of their characterization for holistic assessments and management approaches.}
\textcolor{blue}{An additional objective emerged while addressing the main objective of this
manuscript: sharing information about the distribution and intensity of drivers of change in the St.Lawrence.}
We achieve ~~these objectives~~ \textcolor{blue}{this} with the development of an
open knowledge platform,~~named~~ *eDrivers*,~~. The platform was~~ designed to
facilitate collaboration, real-time assessments of cumulative exposure and to
evolve ~~with the addition of information and threats to~~
\textcolor{blue}{as knowledge accumulates and new threats emerge in}
the St-Lawrence ecosystems.
\textcolor{blue}{Here, we present its guidng principles and accompanying tools.}

@@

## ~~Estuary and Gulf of St. Lawrence~~ \textcolor{blue}{St. Lawrence System}

The St. Lawrence System is formed by the \textcolor{blue}{St. Lawrence} Estuary and the Gulf of St. Lawrence
(Figure \ref{egsl}). The Estuary is defined by the seawater limit
close to Îles d'Orléans to the west and by its connexion to the Gulf
near Pointe-des-Monts.
Estuarine circulation characterizes the Estuary. The surface layer is
composed of a freshwater seaward outflow primarily from the Great Lakes Basin
and through the St. Lawrence River, upstream of the Estuary. Atlantic waters are
flowing mainly from the Cabot Strait landward through the Laurentian Channel
(see below).

@@

\textcolor{red}{\sout{The list of drivers for which we sought data was informed by}} \textcolor{blue}{Drivers selection was informed by} a global
cumulative impact assessment initiative [@halpern2008a; @halpern2015a; @halpern2019] \textcolor{blue}{and available from the National Center for Ecological Analysis and Synthesis (NCEAS) online data repository (Halpern et al. 2015)},
regional holistic evaluations of the state of the St. Lawrence [@dufour2007;
@benoit2012], and communications with regional experts (Table 1).
\textcolor{red}{\sout{We integrated datasets from regional experts and also use global data from the global cumulative impact assessment initiative [@halpern2008a; @halpern2015a]  available from the National Center for Ecological Analysis and Synthesis (NCEAS) online data repository [Table 1; @halpern2015].}}
~~We selected global data that were
unavailable at the regional scale and that were available at a resolution
adequate for use at the scale of the St. Lawrence (*e.g.* marine pollution).~~
\textcolor{blue}{Where regional data on drivers were unavailable, available global data at a resolution adequate for the scale of the St. Lawrence System were used instead (*e.g.* marine pollution).}

We characterized the intensity and distribution of 22 drivers
(Table 1). Drivers incorporated in the analyses are varied in origin,
*i.e.* from terrestrial (*e.g.* nutrient input) to marine (*e.g.* shipping),
and from large scale biophysical processes (*e.g.* temperature anomalies) to
localized anthropogenic activities (*e.g.* fisheries). Drivers were divided
into 4 groups: coastal, climate, fisheries and marine traffic (Table 1). All
data layers and methodologies are described in the supplementary materials.

~~Drivers with non-normal frequency distributions were log-transformed, as in
Halpern et al. 2008, to minimize the impact of extreme outliers that may or may
not be real observations (Figure S1), and all drivers were scaled between 0 and
1 to allow driver comparisons. The 99th quantile of individual driver
distribution was used as the upper bound for scaling.~~
\textcolor{blue}{As in Halpern et al. 2019, drivers with non-normal frequency distributions were log-transformed
to avoid underestimating intermediate driver values. All drivers were also scaled
between 0 and 1 to allow driver comparisons. The 99th quantile of individual
driver distribution was used as the upper bound for scaling to control for
extreme values that may or may not be real observations.}
~~All drivers were embedded in a regular grid composed of 245604 $1 km^2$
hexagonal cells to construct the integrated dataset used for the analyses
(Figure S2).~~
\textcolor{blue}{The St. Lawrence System was divided into a regular grid of
1 $km^2$ hexagonal cells in which all drivers were integrated (Figure S2)}

@@

Hypoxia is mainly distributed in the Laurentian, Anticosti and Esquiman Channels, with the head of the Channels most exposed (Figure \ref{kernel}A). Demersal destructive fisheries are located along the Laurentian Channel, the heads of the Anticosti and Esquiman Channels and around the Magdalen Islands (Figure \ref{kernel}B). By combining both drivers, we can observe that hypoxia and demersal destructive fisheries co-occur mostly at high relative intensity (Figure \ref{kernel}D) in the vicinity of the Anticosti Gyre and the heads of the Esquiman and Anticosti Channels (Figure \ref{kernel}C~~, Box 1~~).

@@

As such, we are launching *eDrivers*, an open knowledge platform focused on
sharing knowledge on the distribution and intensity of drivers and on gathering
a community of experts committed to structuring, standardizing and sharing
knowledge on drivers in support of science and management. In launching this
initiative, our objective is to uphold the highest existing standards of data
management and open science. We identified four guiding principles \textcolor{blue}{(Sections 5.1 to 5.4)} to meet this
objective and that guide the structure of the platform (Figure \ref{diag}).

@@

Through *eDrivers*, these observations will be iteratively improved towards an increasingly robust assessment of cumulative exposure and threat complexes as gaps in knowledge are addressed or approaches to describe drivers are refined. Arguably, the most meaningful benefit anticipated from *eDrivers* will be the gain in efficient access to comparable data-based knowledge on the exposure of ecosystems to multiple threats. \textcolor{blue}{We provide an example of the ease with which this can be accomplished using the \textit{eDrivers} R package to reproduce figure \ref{kernel} (Box 1).} This could pay quick scientific and management dividends by \textcolor{blue}{efficiently} drawing on the knowledge and efforts of a wide range of contributors, by expanding avenues of scientific inquiry, by decreasing overall effort duplication and research costs, and by increasing research efficiency [@franzoni2014].
